### 2014-10-21 Paper_v1.tex
Abstract and introduction

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

{\noindent \textbf{Keywords:}~Coagulation, Mathematical modeling, Systems Biology}

@@

%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
%Many have suggested that the integration of experimental and computational research is required to unravel
%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
%like thrombosis during surgery.
%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
%Development and validation of detailed network models requires a significant investment of time and material resources. 
%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.

@@

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, but the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics. 
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna and coworkers showed that model performance is typically 
sensitive to only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions from sloppy models may be possible, despite parametric uncertainty, if a few critical components are well understood. 
Perhaps a more important issue is that ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Data-driven \citep{Cirit:2012kx} or logical model approaches \citep{Morris:2011ys} 
are emerging modeling paradigms that constrain model structure by the availability of experimental data. 
These are interesting techniques because the data naturally informs model structure, thereby limiting the 
identification challenge. Although not always mechanistic, these models can simulate surprisingly rich system responses in the absence of mechanism [REFHERE]. 
However, modeling dynamics using logical models is not straightforward. 
Thus, a potential third approach could be a combination of ODEs and logical models, where mechanistic information was used when available, 
and missing or incomplete knowledge encoded using causal local rules.

In this study, we developed a hybrid approach which combined ODEs and logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

Coagulation models in the literature have typically been formulated as systems of nonlinear ordinary differential equations, 
often using mass-action kinetics to describe the rates of biochemical conversions [REFHERE].
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\section*{Discussion}
In this study, we developed a reduced order model of the human coagulation cascade.
We chose coagulation because it is well studied, has a complex architecture, 
and has an abundance of experimental data available for model identification and validation. 
However, coagulation was just a proof-of concept; 
the proposed framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems. 
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
However, not all of the model parameters were uniquely identifiable, given the training data.
Instead, we estimated an ensemble of likely parameter sets (N = 20)
from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
be used to model other biochemical networks important to human health.

Malfunctions in coagulation can have potentially fatal consequences.
For example, aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
Coagulation management during surgery can also be challenging, particularly because of the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in-vivo} \cite{TOMOKIYO2003}.
Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
However, rFVIIa requires frequent administration (every 2-3 hr) because of its short half-life in the circulation. 
Many questions also remain about the mechanism of action of rFVIIa, its effective dose range \cite{TOMOKIYO2003}, and its overall utility 
for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. In the current proof of concept model, we did not model rFVIIa directly. 
Rather we modeled a broad class of trigger materials which initiated the extrinsic coagulation cascade. Since we identified the model using TF/FVIIa, inherent to
our rFVIIa simulations was the presence of TF. However, even with this complication, 
the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
Addition of rFVIIa directly activated thrombin through the fluid phase initiation pathway. 
However, in the presence of ATIII and the protein C pathway but without fVIII or fIX, 
this initial thrombin signal was quickly extinguished by inhibition, as thrombin amplification could not occur. 
As the dose of rFVIIa increased, the peak thrombin time decreased, eventually saturating.
Butenas \emph{et al.,} performed an extensive study of rFVIIa induced thrombin generation in hemophilia \citep{Butenas:2002ab}.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
\caption{Reduced order coagulation model predictions of rFVIIa administration. 
Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 50x) in the absence of fVIII and fIX.
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
\end{figure}

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Dynamic Modeling of the Human Coagulation Cascade using Reduced Order Effective Kinetic Models}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling blood coagulation}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Processes}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
In this study, we present a novel hybrid modeling approach which combines ordinary differential equation (ODE) modeling with logical rules to simulate an archetype biochemical network, 
the human coagulation cascade.  
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
This formulation was more than an order of magnitude smaller than current coagulation models, 
because many of the mechanistic details of coagulation were encoded as logical rules.
We estimated an ensemble of likely model parameters (N = 20) from \textit{in~vitro} extrinsic coagulation data sets, with and without inhibitors, 
by minimizing the residual between model simulations and experimental measurements using particle swarm optimization (PSO). 
We then validated the model ensemble using thrombin data sets that were not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the hybrid approach produced a surprisingly predictive model given its small size, 
suggesting the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems.

\vspace{0.25in}
{\noindent \textbf{Keywords:}~Blood coagulation, Mathematical modeling, Systems biology}

@@

%We chose coagulation because it is well studied, has a complex architecture, 
%and has an abundance of experimental data available for model identification and validation. 
%However, coagulation was just a proof-of concept to validate our approach; 
%

%Developing mathematical modeling tools to interrogate the molecular basis of human disease is one of the grand challenges facing systems biology.
%Developing computational tools to understand the molecular basis of human disease is a grand challenge facing systems biology.
%Many have suggested that the integration of experimental and computational research is required to unravel
%critical questions facing molecular medicine \cite{KITANO2002,ASSMUS2006,ARNAUD2006}. 
%Despite this, mechanistic network modeling has not played a significant role in the development of new therapies for cancer, cardiovascular diseases or for acute events  
%like thrombosis during surgery.
%A critical limiting issue often cited has been model uncertainty \cite{BAILEY2001}. 
%Development and validation of detailed network models requires a significant investment of time and material resources. 
%Alternatively, development of tools using toy models often fails to demonstrate relevance to biologists or physicians.  
%What is needed is a model system, that is simultaneously tractable and directly applicable to the study of disease.
%Human blood is perhaps such an ideal model system \cite{Diamond:2009yf}. 
%Blood is a complex mixture of proteins and cells which collectively carry out many functions.
%However, one well studied facet of blood, which could serve as a valuable model system, is the coagulation cascade. 
%Understanding the mechanism of rFVIIa might provide insight into the therapeutic dose under different physiological conditions.
%More generally, rFVIIa treatment could also serve an ideal test case to develop strategies which retrospectively estimate operational perturbations associated with drugs.
%For example, the effect of rFVIIa is directly observable in coagulation outputs such as thrombin concentration.
%Thus, by analyzing these outputs in a normal versus a perturbed network, it might be possible to reverse engineer the molecular influence associated with a therapeutic intervention.

@@

\section*{Introduction}
Developing mathematical models of biochemical networks is a significant facet of systems biology.
Modeling approaches differ in their degree of detail, where the choice of approach is often determined by prior knowledge, or model requirements \citep{2012_kholodenko_kolch_SciSig}. 
Ordinary differential equation (ODE) models are a common tool for modeling biochemical systems because of their ability to capture dynamics, and encode mechanism.
However, ODE models typically come with difficult (or sometimes impossible) parameter identification problems. 
For example, Gadkar et al., showed that even with near-perfect information, 
it was often impossible to identify all the parameters in typical signal transduction models \citep{Gadkar:2005ad}. 
However, it is not clear whether we actually need precise estimates for all model parameters. 
Bailey suggested more than a decade ago, that achieving qualitative or even quantitative understanding of biological systems 
should not require complete structural and parametric knowledge \citep{2001_bailey_NatBiotech}. 
Since Bailey's complex biology with no parameters hypothesis, Sethna showed that model performance is typically 
sensitive to only a few parameters, a characteristic seemingly universal to multi-parameter models referred to as \textit{sloppiness} \citep{Machta:2013aa}.
Thus, reasonable predictions may be possible, despite parametric uncertainty, if a few critical parameters are well defined. 
Perhaps more importantly, ODE models also require significant mechanistic information, thereby limiting their utility in poorly understood systems.
Toward this challenge, logical modeling is an emerging paradigm that encodes causal relationships using non-linear transfer functions \citep{Morris:2010aa}.
These functions encode interactions between model components, without requiring mechanism.
Despite their simplicity, logical models have captured rich system behaviors in the absence of mechanistic information \citep{Saez-Rodriguez:2011aa,Morris:2011ys,Morris:2012aa}. 
However, modeling complex dynamics remains difficult with this class of model.
Thus, there is an unmet need for a third approach which combines ODEs and logical models, where ODEs could encode mechanistic information, 
while missing or incomplete mechanistic knowledge can be approximated using a logical approach.

In this study, we developed a hybrid approach which combines ODE modeling with logical rules to model a well studied biochemical network, 
the human coagulation system. Coagulation is an archetype proteolytic cascade involving both positive 
and negative feedback \cite{Butenas:2002aa,Schenone:2004aa,Adams:2009aa}. 
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets. 
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination \cite{GOLDHABER2006, Brummel:2002aa}.
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal. 
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood \cite{MANN1990,ROBERTS1998,MANN1999}.
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by FVIIa \cite{Butenas:2002aa}. 
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, 
leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin. 
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. 
Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII)
neutralizes several of the proteases generated during coagulation, including thrombin.  
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR),
converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa. 
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as activated protein C (APC) or antithrombin III (ATIII).

Coagulation models in the literature are typically formulated as systems of nonlinear ordinary differential equations, 
using mass-action or other kinetics, to describe the rates of biochemical conversions \citep{Kuharsky:2001aa}.
Previous ODE coagulation models from our laboratory \citep{Luan:2007aa,2010_luan_varner_MolBioSys}, 
were built upon the earlier studies of Jones and Mann \citep{Jones:1994aa}, 
and later Hockin \emph{et al.} \citep{Hockin:2002aa}, who developed and then refined highly mechanistic models of extrinsic coagulation. 
Other laboratories have also expanded upon Hockin \emph{et al.}, for example including the intrinsic pathway, and thrombin mediated clot formation \citep{Chatterjee:2010aa}. 
Other biochemical or biophysical aspects of coagulation have also been modeled, such as platelet biochemistry \citep{Stalker:2013aa}, 
multi-scale models of clot formation \citep{Leiderman:2014aa, Bannish:2014ab}, and transport inside growing clots \citep{Voronov:2013aa}.
However, most of these previous studies were based upon detailed mechanistic knowledge. 
Such a description is possible because blood, while enormously complex, can be systematically interrogated. 
However, other systems, such as the intracellular networks which mediate cancer, are more difficult to experimentally probe and model.  
Towards this unmet need, we have formulated a hybrid modeling approach which combines ODEs and logical rules to model 
biochemical processes for which mechanistic information is difficult or impossible to obtain. 
We have used this approach to model the coagulation cascade because of the wealth of training and validation data available for this system. 
We proposed a reduced order coagulation model that consisted of five differential equations augmented with several logical rules.
We identified the reduced order coagulation model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
We then validated the model using thrombin data sets from multiple laboratories, for both normal and hemophilic coagulation.
Flux and sensitivity analysis identified which parameter or parameters combinations were critical to model performance in several conditions. 
The reduced order hybrid approach produced a surprisingly predictive coagulation model, suggesting this framework could potentially be used to model other biochemical networks important to human health.

@@

\section*{Discussion}
In this study, we developed a reduced order model of the human coagulation cascade.
We chose coagulation because it is well studied, has a complex architecture, 
and has an abundance of experimental data available for model identification and validation. 
However, coagulation was just a proof-of concept validation of our approach; 
the proposed hybrid framework could also be used to dynamically model other biochemical networks, including intracellular metabolic networks, gene expression programs 
or potentially even cell free metabolic systems. 
The model consisted of five differential equations augmented with several logical rules describing 
regulatory connections between model components, and unmodeled interactions in the network. 
We estimated model parameters from in vitro extrinsic coagulation data sets, in the presence of ATIII, with and without the protein C pathway. 
To estimate parameters, the residual between model simulations and experimental measurements was minimized using particle swarm optimization (PSO). 
However, not all of the model parameters were uniquely identifiable, given the training data.
Instead, we estimated an ensemble of likely parameter sets (N = 20)
from eight \emph{in vitro} time-series coagulation data sets with and without the protein C pathway.
We validated the ensemble using thrombin data sets taken from multiple laboratories for a variety of experimental conditions not used during training. 
The ensemble predicted thrombin trajectories for conditions not used for model training, including thrombin generation for normal and hemophilic coagulation 
in the presence of platelets (a significant unmodeled component).
We then used flux analysis to understand how the network operated in a variety of conditions, and global sensitivity analysis to identify which parameters controlled the 
performance of the network. Taken together, the reduced order hybrid approach produced a surprisingly predictive model, suggesting this approach could potentially 
be used to model other biochemical networks important to human health.

Malfunctions in coagulation can have potentially fatal consequences.
Aggressive clotting involved with Coronary Artery Diseases (CADs), collectively accounts for 38\% of all deaths in North America \cite{HANSSON2005}.
Coagulation management during surgery can also be challenging, particularly with the increasing clinical use of antithrombotic drugs \cite{Tanaka:2009wo}.   
Insufficient coagulation due to genetic disorders such as hemophilia can also result in recurrent bleeding. 
The coagulation factors VIII (fVIII) and IX (fIX) are deficient in Hemophilia A and B, respectively \cite{COOPER1994,MANNUCCI2001,MITCHELL2008}.
People with mild hemophilia have 5-40\% of the normal clotting factor levels while severe hemophiliacs have $<$1\% \cite{MITCHELL2008}.
Hemophilia can be controlled with regular infusions of the deficient clotting factors. 
However, clotting factor replacement sometimes leads to the formation of fVIII and fIX inhibitors \emph{in-vivo} \cite{TOMOKIYO2003}.
Alternatively, recombinant factor VIIa (rFVIIa) has been used to treat bleeding disorders \cite{Hedner:2008rm,Talbot:2009kk} including
hemophilia with and without factor VIII/IX inhibitors \cite{Shapiro:2008px}.
However, rFVIIa requires frequent administration (every 2-3 hr), and many questions remain about its mechanism of action, 
its effective dose range \cite{TOMOKIYO2003}, and its overall utility for the treatment trauma-associated hemorrhage \cite{Duchesne:2008hw}. 
In this study, we did not model rFVIIa induced coagulation directly. 
Rather, we modeled a general trigger which initiated the extrinsic coagulation cascade. 
Since we identified the model using TF/FVIIa, inherent to our rFVIIa simulations (and the rate constant governing initiation) 
was the presence of TF. However, even with this complication, 
the model generated potentially useful insight into the rFVIIa mechanism of action, and its possible shortcomings especially for the treatment of hemophilia. 
The addition of rFVIIa directly activated thrombin through the initiation pathway. 
However, no amplification of the thrombin signal occurred without fVIII or fIX. 
Thus, the peak thrombin signal was lower than normal coagulation, the peak thrombin time was longer, and the thrombin signal was eventually inhibited by ATIII and the protein C pathway.
However, as the dose of rFVIIa increased, the peak thrombin time decreased (eventually saturating around 100$\times$nominal trigger), and the peak thrombin value increased.
For 100$\times$nominal trigger, the thrombin profile resembled normal coagulation.
Butenas \emph{et al.,} performed an extensive in vitro study of rFVIIa induced thrombin generation under normal and hemophilic conditions \citep{Butenas:2002ab}.
They found qualitatively similar trends, namely rFVIIa could restore normal coagulation (even in the absence of TF) for large enough rFVIIa doses, although rFVIIa induced coagulation 
in hemophilia (even for large rFVIIa doses) lagged the normal profile. 
These results suggest that rFVIIa administration alone
might not be able to initiate normal coagulation in recurrent bleeding, unless the dosage is above a critical threshold. 
Unfortunately, this threshold is likely patient specific, thus limiting potentially limiting the utility of rFVIIa administration.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-7-rFVIIa-Trigger.pdf}
\caption{Reduced order coagulation model predictions of rFVIIa administration. 
A: Simulations of thrombin formation in the presence of ATIII and the protein C pathway were conducted for a range of trigger values (1x - 200x nominal) in the absence of fVIII and fIX.
B: Comparison of thrombin generation for normal versus hemophilia for 10x nominal trigger.
Solid lines denote the simulated mean value of the thrombin profile for N = 20 independent particles.
The shaded region denotes the 99\% confidence estimate of the mean simulated thrombin value (uncertainty in the model simulation). 
All factors and control proteins were at their physiological concentration unless others noted.}\label{fig-rFVII}
\end{figure}
